---
title: "6372 - Project 1 (Josh Initial)"
author: "Josh"
date: "2026-05-21"
output: html_document
editor_options: 
  chunk_output_type: console
---

# Objective 1
Hospitals are constantly trying to understand what impacts the duration of a
patients stay. The data set consists of 11 variables collected on each of 113 hospitals. One of
the variables is the hospitals infection risk, which is an estimated probability of a patient
acquiring a hospital related infection. The hospitals are interest to know if the infection risk
and related factors are associated with a patient’s length of stay even after accounting for other
variables that would/may impact hospital stays such as Age, Number of nurses, number of
patients per day, number of beds, routine medical checks, etc. Produce a regression model to
address the primary question.


# Libraries
```{r}
library(car)
library(corrplot)
library(GGally)
library(tidyverse)
library(dplyr)
library(caret)   #for prediction models and k-fold CV
library(naniar)  #for graphing missing data behavior
```

# Read in Data
```{r}
df = read.csv('/Users/joshtsamuel/SMU/6372 Applied Statistics- Inference and Modeling/Project 1/HospitalDurations.csv')
```

```{r}
str(df)
vis_miss(df)
dim(df)
# no missing values
```

```{r}
df = df %>%
  mutate(across(c(Med.Sc.Aff, Region), as.factor))
str(df)
```

# EDA

```{r}
response = "Lgth.of.Sty"
predictors = setdiff(names(df), c(response, "ID"))
ggpairs(df,columns = c(response, predictors))

# Region 1 - 4 seems to go in order as longest length of stay to smallest

ggplot(df, aes(x = Region, y = Lgth.of.Sty)) +
  geom_boxplot() +
  labs(
    x = "Region",
    y = "Length of Stay",
    title = "Length of Stay by Region"
  )

ggplot(df, aes(x = Med.Sc.Aff, y = Lgth.of.Sty)) +
  geom_boxplot() +
  labs(
    x = "Med School Affiliation (1 = Yes, 2 = No)",
    y = "Length of Stay",
    title = "Length of Stay by Medical School Affiliation"
  )

# hospitals who do have a medical school affiliation have avg longer length of stay

ggplot(df, aes(x = Inf.Risk, y = Lgth.of.Sty, color = Region)) +
  geom_point() + geom_smooth()
  labs(
    x = "Infection Risk",
    y = "Length of Stay",
    title = "Infection Risk vs Length of Stay By Region "
  )
```

```{r}
df_num <- df %>%
  select(where(is.numeric)) %>%
  select(-ID)   # remove ID column if needed

cor_matrix <- cor(df_num, use = "complete.obs")
cor_matrix

install.packages("corrplot")
library(corrplot)

corrplot(
  cor_matrix,
  method = "color",
  type = "upper",
  addCoef.col = "black",
  tl.cex = 0.8,
  number.cex = 0.7
)

## Insights
# n.beds -> avg.pat & avg.nur VERY HIGH CORR & pct.ser.fac HIGH CORR
# avg.pat -> n.beds VERY HIGH CORR & pct.ser.fac HIGH CORR
# avg.nur -> pct.ser.fac HIGH CORR
# inf.risk -> R.cul.rat MID CORR & R.CX.ray.Rat MED CORR
# inf.risk -> kinda MED CORR to all
```

# Modeling


```{r}
# normal
mlr1 = lm(Lgth.of.Sty ~ . - ID, data = df)
summary(mlr1)

# Assumptions
par(mfrow = c(2, 2))
plot(mlr1)
par(mfrow = c(1, 1))

# Checking for Multicollinearity
vif(mlr1)

# With Interaction
mlr2 = lm(Lgth.of.Sty ~ . - ID + Inf.Risk*Region, data = df)
summary(mlr2)

# Assumptions
par(mfrow = c(2, 2))
plot(mlr2)
par(mfrow = c(1, 1))

# log transform
mlr3 = lm(log(Lgth.of.Sty) ~ . - ID, data = df)
summary(mlr3)

# Assumptions
par(mfrow = c(2, 2))
plot(mlr3)
par(mfrow = c(1, 1))




```



